Assignment 3

### 2019-08-07 ex2/report/report.tex
Fix typos if Ex2 and small comments

@@

For research the invention and assessment of techniques for network anomaly detection poses many challenges. To implement such a system, a considerable number of features can be extracted from network data which might be beneficial for anomaly detection and it is not always clear which features are actually important. For the training of models usually datasets are used which have been generated artificially in a controlled test environment.

As a downside of this approach, it is unclear whether a machine learning model learns to classify based on characteristics that are inherent to the attacks which should be detected, or rather learns to classify based on patterns that were unintentionally created during dataset generation.

@@

\subsection{Individual Conditional Expectation}
The averaging which is done for PD plots introduces the problem that the influence of a feature on individual samples is lost. In our case, the predictions of both deep learning and random forests turned out to be strongly dependent on other features as well. Hence, by averaging information gets lost and wrong explanations might be the consequence.

@@

\begin{equation}
\frac{d}{dy} \text{ALE}_i (y) = \mathbb E_{\boldsymbol X | X_i}\left(\frac{d}{dy} f(X_1,\ldots,X_{i-1},y,X_{i+1},\ldots X_n) | X_i=y\right)
\end{equation}

@@

We generated PD plots for all features contained in our extracted feature vector. Fig.~\ref{fig:pdp} shows examples for some features for the different folds. Comparing plots generated for our deep learning and for our random forest approach, respectively, the two machine learning approaches show the outstanding difference that deep learning PDPs are smooth, while random forest PDPs fluctuate heavily. Considering the functioning of neural networks and decision trees, this is an expected result.

@@

Plots depicted in Fig.~\ref{fig:ale} show a certain similarity to the PD plots in Fig.\ref{fig:pdp}. However, as apparant at the \textit{srcPort}, explanations from PD and ALE plots might also deviate substantially.

An important observation for ALE plots particularly for random forests is that influce of TTL values is even more distinctive. In fact, judging from ALE plots the random forest seems to almost exclusively consider the TTL for detecting attacks. This emphasizes the problem described already for PDPs.

@@

\caption{Examples for ICE plots for deep learning (left side) and random forest (right side).}
\label{fig:ice}
\end{figure*}

@@

The fact that results for the training process. with predictions are very similar to original data is due to the good detection performance of random forests. Hence, there is hardly any difference when training with the real labels and with the prediction of the random forest. It is interesting to note that training with predictions from deep learning even results in slightly better performance than training with real labels.

@@

Hence, for this task we first had to find a pattern usable as a backdoor in feature vectors. On the one hand, this pattern has to be detectable reliably by the anomaly detection system, one the other hand it must be possible to generate real network traffic which translate into feature vectors exhibiting these patterns.

>>>>

\newcommand\note[2]{{\color{#1}#2}}
\newcommand\todo[1]{{\note{red}{TODO: #1}}}

@@

For research, the invention and assessment of techniques for network anomaly detection poses many challenges. To implement such a system, a considerable number of features can be extracted from network data which might be beneficial for anomaly detection and it is not always clear which features are actually important.

For the training of models usually datasets are used which have been generated artificially in a controlled test environment. As a downside of this approach, it is unclear whether a machine learning model learns to classify based on characteristics that are inherent to the attacks which should be detected, or rather learns to classify based on patterns that were unintentionally created during dataset generation.

@@

\subsection{Individual Conditional Expectation}
The averaging which is done for PD plots introduces the problem that the influence of a feature on individual samples is lost. In our case, the predictions of both deep learning and random forests turned out to be strongly dependent on other features as well \todo{How do we see that?}. Hence, by averaging information gets lost and wrong explanations might be the consequence.

@@

\begin{equation}
\frac{d}{dy} \text{ALE}_i (y) = \mathbb E_{\boldsymbol X | X_i}\left(\frac{d}{dy} f(X_1,\ldots,X_{i-1},y,X_{i+1},\ldots X_n) | X_i=y\right)
\end{equation}
\todo{I think that there are some minor problems with $y$ and $X_i$ in the formula.}

@@

We generated PD plots for all features contained in our extracted feature vector. Fig.~\ref{fig:pdp} shows examples for some features for the different folds. Comparing plots generated for our deep learning and for our random forest approach, respectively, the two machine learning approaches show the difference that deep learning PDPs are smooth, while random forest PDPs fluctuate heavily. Considering the functioning of neural networks and decision trees, this is an expected result.

@@

Plots depicted in Fig.~\ref{fig:ale} show a certain similarity to the PD plots in Fig.\ref{fig:pdp}. However, as apparant at the \textit{srcPort}, explanations from PD and ALE plots might also deviate substantially \todo{Are we sure that this is not a bug?}.

An important observation for ALE plots particularly for random forests is that influence of TTL values is even more distinctive. In fact, judging from ALE plots the random forest seems to almost exclusively consider the TTL for detecting attacks. This emphasizes the problem described already for PDPs.

@@

\caption{Examples for ICE plots for deep learning (left side) and random forest (right side) \todo{Fixme: No title!}.}
\label{fig:ice}
\end{figure*}

@@

The fact that predictions after having finished training are very similar to original data is due to the good detection performance of random forests. Hence, there is hardly any difference when training with the real labels and with the prediction of the random forest. It is interesting to note that training with predictions from deep learning even results in slightly better performance than training with real labels.

@@

Hence, for this task we first had to find a pattern usable as a backdoor in feature vectors. On one hand, this pattern has to be detectable reliably by the anomaly detection system, one the other hand it must be possible to generate real network traffic which translate into feature vectors exhibiting these patterns.

### 2019-08-07 ex3/report/report.tex
Fix typos if Ex2 and small comments

\subsection{Feature Importance Metrics}
\begin{table}
\caption{Accuracy drop when replacing feature values with random data.}
\label{tab:feat_importance}
\begin{tabular}{l r}
\toprule
Feature & Accuracy drop \\ \midrule
Protocol	&	0.1323	\\
PacketLength	&	0.1203	\\
SYN Flag	&	0.1067	\\
ACK Flag	&	0.0805	\\
Destination port	&	0.0796	\\
PSH Flag	&	0.0767	\\
Direction	&	0.0732	\\
Source port	&	0.0254	\\
Interarrival time	&	0.0200	\\
RST Flag	&	0.0116	\\
FIN Flag	&	0.0100	\\
ECE	Flag &	0.0000	\\
URG Flag	&	0.0000	\\
CWR Flag	&	0.0000	\\
NS Flag	&	0.0000	\\
\bottomrule
\end{tabular}
\end{table}
As a first step to understanding the neural network's decisions, we tried to estimate how important individual features are for the model's predictions.

@@

\subsection{Explainability Plots}
During flow extraction we used a usual 5-tuple flow key, which distinguishe flows based on the protocol they use and their source and destination port and IP address. Hence, these features are constant for all packets in one flow, which allows us to use explainability methods which have been proposed for non-recurrent classification methods.

@@

\caption{The solid line shows the mean value of the feature for all flows of a specific attack type. The shaded region shows the change in confidence that occurs when the feature is varied.}
\label{fig:plot_features}
\end{figure*}

>>>>

\subsection{Feature Importance Metrics}
\begin{table}
\caption{Accuracy drop when replacing feature values with random data.}
\label{tab:feat_importance}
\begin{tabular}{l r}
\toprule
Feature & Accuracy drop \\ \midrule
Protocol	&	0.1323	\\
Packet Length	&	0.1203	\\
SYN Flag	&	0.1067	\\
ACK Flag	&	0.0805	\\
Destination port	&	0.0796	\\
PSH Flag	&	0.0767	\\
Direction	&	0.0732	\\
Source port	&	0.0254	\\
Interarrival time	&	0.0200	\\
RST Flag	&	0.0116	\\
FIN Flag	&	0.0100	\\
ECE	Flag &	0.0000	\\
URG Flag	&	0.0000	\\
CWR Flag	&	0.0000	\\
NS Flag	&	0.0000	\\
\bottomrule
\end{tabular}
\end{table}
As a first step to understanding the neural network's decisions, we tried to estimate how important individual features are for the model's predictions.

@@

\subsection{Explainability Plots}
During flow extraction we used a usual 5-tuple flow key, which distinguishes flows based on the protocol they use and their source and destination port and IP address. Hence, these features are constant for all packets in one flow, which allows us to use explainability methods which have been proposed for non-recurrent classification methods.

@@

\caption{The solid line shows the median value of the feature for all flows of a specific attack type. The shaded region shows the change in confidence that occurs when the feature is varied.}
\label{fig:plot_features}
\end{figure*}
